---
title: "Size Comparisons: Dataset-JSON vs. XPT"
---

## Overview

This report examines one of the practical considerations when evaluating Dataset-JSON as a replacement for the legacy SAS XPORT (XPT) format: **file size**. The XPT format allocates fixed-width fields and pads records with blanks, which can waste considerable storage. Dataset-JSON, being a text-based format, compresses variable-length data efficiently for sparse or short-string heavy datasets — but can be larger for numerically dense data due to JSON's human-readable representation.

The datasets used here come from a real clinical trial study (CDISC SDTM pilot data), spanning domain types such as demography, adverse events, laboratory results, questionnaires, and supplemental qualifiers. By comparing file sizes side-by-side we can see where Dataset-JSON offers storage savings and where XPT may be more compact.

**Key questions this analysis answers:**

- How does the size of each Dataset-JSON file compare to its XPT counterpart?
- Which datasets benefit most from the JSON format?
- Are there datasets where the JSON representation is actually larger than XPT?

---

```{r}
#| echo: false
#| message: false
#| warning: false

library(dplyr)
library(tidyr)
library(DT)
library(plotly)
library(ggplot2)

# ── Read file sizes ─────────────────────────────────────────────────────────

xpt_files  <- list.files("xpts/", pattern = "\\.xpt$",  full.names = TRUE)
json_files <- list.files("json/",  pattern = "\\.json$", full.names = TRUE)

xpt_sizes  <- setNames(file.info(xpt_files)$size,
                       tools::file_path_sans_ext(basename(xpt_files)))
json_sizes <- setNames(file.info(json_files)$size,
                       tools::file_path_sans_ext(basename(json_files)))

# Exclude ts (no JSON generated for it)
common_names <- intersect(names(xpt_sizes), names(json_sizes))

size_df <- tibble(
  Dataset = toupper(common_names),
  XPT_KB  = round(xpt_sizes[common_names]  / 1024, 1),
  JSON_KB = round(json_sizes[common_names] / 1024, 1)
) |>
  mutate(
    Ratio         = round(JSON_KB / XPT_KB, 3),
    `Size Change` = case_when(
      Ratio < 1  ~ "JSON smaller",
      Ratio > 1  ~ "JSON larger",
      TRUE       ~ "Equal"
    )
  ) |>
  arrange(desc(XPT_KB))
```

## Interactive Size Comparison Table

The table below shows all `r nrow(size_df)` datasets. By default it displays **12 rows** sorted by XPT file size (largest first). You can:

- **Sort** any column by clicking the column header
- **Search** across all columns using the search box
- **Adjust** the number of rows shown using the "Show entries" control
- **Page** through results using the navigation at the bottom

```{r}
#| echo: false

display_df <- size_df |>
  select(
    Dataset,
    `XPT Size (KB)`  = XPT_KB,
    `JSON Size (KB)` = JSON_KB,
    `JSON / XPT Ratio` = Ratio,
    `Size Change`
  )

datatable(
  display_df,
  options = list(
    pageLength    = 12,
    lengthMenu    = list(c(5, 10, 12, 15, 20, -1),
                         c("5", "10", "12", "15", "20", "All")),
    order         = list(list(1, "desc")),   # default sort: XPT Size desc
    scrollX       = TRUE,
    dom           = "lfrtip"
  ),
  rownames = FALSE,
  class    = "stripe hover cell-border compact",
  filter   = "top"
) |>
  formatStyle(
    "Size Change",
    backgroundColor = styleEqual(
      c("JSON smaller", "JSON larger", "Equal"),
      c("#d4edda",      "#f8d7da",     "#fff3cd")
    )
  ) |>
  formatStyle(
    "JSON / XPT Ratio",
    background = styleColorBar(range(display_df[["JSON / XPT Ratio"]]), "#c8e6fb"),
    backgroundSize   = "100% 90%",
    backgroundRepeat = "no-repeat",
    backgroundPosition = "center"
  )
```

---

## Bar Chart: XPT vs. JSON File Sizes (Top 12 by XPT Size)

The chart below focuses on the 12 largest datasets (by XPT size). Hover over any bar for exact values.

```{r}
#| echo: false
#| fig-height: 7

top12 <- size_df |> slice(1:12)

long12 <- top12 |>
  select(Dataset, `XPT (KB)` = XPT_KB, `JSON (KB)` = JSON_KB) |>
  pivot_longer(cols = c(`XPT (KB)`, `JSON (KB)`),
               names_to  = "Format",
               values_to = "Size_KB") |>
  mutate(Dataset = factor(Dataset, levels = rev(top12$Dataset)))

p <- ggplot(long12, aes(x = Dataset, y = Size_KB, fill = Format,
                        text = paste0(Format, ": ", Size_KB, " KB"))) +
  geom_col(position = "dodge") +
  coord_flip() +
  scale_fill_manual(values = c("XPT (KB)" = "#00BFC4", "JSON (KB)" = "#F8766D")) +
  labs(
    x    = NULL,
    y    = "File Size (KB)",
    fill = "Format"
  ) +
  theme_minimal(base_size = 13) +
  theme(legend.position = "top")

ggplotly(p, tooltip = "text") |>
  layout(legend = list(orientation = "h", x = 0, y = 1.08))
```

---

## Ratio Chart: JSON / XPT File Size Ratio

A ratio below **1.0** (dashed red line) indicates the JSON file is **smaller** than the XPT file; a ratio above 1.0 means JSON is larger. Hover for dataset details.

```{r}
#| echo: false
#| fig-height: 7

ratio_df <- size_df |>
  arrange(Ratio) |>
  mutate(Dataset = factor(Dataset, levels = Dataset))

p2 <- ggplot(ratio_df,
             aes(x = Dataset, y = Ratio,
                 fill = `Size Change`,
                 text = paste0(Dataset, "\nRatio: ", Ratio,
                               "\nXPT: ", XPT_KB, " KB",
                               "\nJSON: ", JSON_KB, " KB"))) +
  geom_col() +
  geom_hline(yintercept = 1, linetype = "dashed", color = "red", linewidth = 0.8) +
  coord_flip() +
  scale_fill_manual(values = c("JSON smaller" = "#28a745",
                                "JSON larger"  = "#dc3545",
                                "Equal"        = "#ffc107")) +
  labs(
    x    = NULL,
    y    = "JSON Size / XPT Size",
    fill = NULL
  ) +
  theme_minimal(base_size = 13) +
  theme(legend.position = "top")

ggplotly(p2, tooltip = "text") |>
  layout(legend = list(orientation = "h", x = 0, y = 1.08))
```

---

## Summary

```{r}
#| echo: false

smaller_count <- sum(size_df$Ratio < 1)
larger_count  <- sum(size_df$Ratio > 1)
avg_ratio     <- round(mean(size_df$Ratio), 2)
max_saving    <- size_df |> filter(Ratio == min(Ratio)) |> pull(Dataset)
max_increase  <- size_df |> filter(Ratio == max(Ratio)) |> pull(Dataset)
```

Across `r nrow(size_df)` datasets compared:

- **`r smaller_count`** datasets are **smaller** as Dataset-JSON than as XPT.
- **`r larger_count`** datasets are **larger** as Dataset-JSON than as XPT.
- The **average JSON/XPT ratio** is **`r avg_ratio`** (values below 1 indicate overall space savings).
- The biggest space saving is in **`r max_saving`**, while **`r max_increase`** sees the largest size increase in JSON format.

These results highlight that Dataset-JSON is not universally more compact than XPT. Datasets with many short fixed-width fields (especially padded numerics) tend to compress well in JSON, while datasets with many long, dense numeric columns may grow. Understanding this trade-off is important when planning a regulatory submission strategy centred on Dataset-JSON.
